Cultural Transmission and Technological Transitions during the Late Paleolithic in Korea

### 2022-11-30 analysis/paper/paper.qmd
editing discussion

@@

Panel A of @fig-plot-cv-all-attributes shows a relatively wide distribution of body length (BL) and maximum length (ML) while the other attributes are more similar and shorter (under 2.5 cm). Among these attributes, stem width (SW) is the smallest, regardless of overall size of the stemmed points. This result may imply that the stem width was the most constrained dimension of the artefact. We might speculate that this is due to the requirements of halfting with wooden shafts, where the shafts were selected from a small range of diameters that optimised strength, weight and flexibility.

@@

@@

To investigate the social contexts of technological transitions represented by the emergence of stemmed points during the Korean Late Paleolithic, we applied concepts of cultural transmission. We asked three questions to examine the dominant mode of cultural transmission and the transmission process over time: what is the dominant mode of cultural transmission for technological innovation in the Korean Late Paleolithic? Do the modes of cultural transmission vary over time? and Do the modes of cultural transmission vary over space? We focused on two transmission biases: guided variation (socially isolated groups made stemmed points through trial and error) and indirect bias (socially connected groups whose knowledge of stemmed points derived from copying others). We set roughly 25 as our threshold of the CV value to decide between these two types of transmission biases, based on prior work on Great Basin Projectiles [@eerkens2001techniques]. If the CV value is lower than 25, which indicates indirect bias while the value is 25 or higher means guided variation.

The CV values for all stemmed points are mostly over 25, except for maximum length. PCA analysis shows positive relationships between attributes (@fig-plot-cv-all-attributes panel B). We then examined the CV values for chronological phases in the Korean Late Paleolithic period (@fig-plot-cv-over-time). After excluding two stemmed points for the Phase 1 to compute CV due to the small sample size issue, the overall CV values become lower. In addition to the maximum length, body length, mid width, and tang width for the Phase 2 are under 25. We observe the different degree of contribution by each attribute to the shape of stemmed points. For example, tang and stem related attributes have higher CV values compared to the body related attributes. We expect that people chose to emulate the body related attributes and had more freedom to produce the tang and stem related attributes.

Compared to the Phase 2 and 3, we found that there are only minor differences between the two phases. The CV of the two most important attributes, body length and maximum length, slightly increased in Phase 3. The increase is not significant to generalize (p \> 0.05, @tbl-cv-over-time), we explored individual assemblages including multiple stemmed points that could show a more clear pattern. The four assemblages in Suyanggae and Yongsandong sites show that CV values decreased with narrower confidence intervals, which means a high degree of precision (@fig-plot-cv-four-assemblage). This could indicate the stronger connections within the groups during Phase 3.

@@

There is no clear pattern variation in CV between the sites. Some sites such as Sachang have very high CV values while other sites including Suyanggae and Jingeunuel have lower CV (@fig-plot-cv-by-site). We assumed that the CV could be impacted by other factors such as the choice of raw materials (@fig-plot-cv-by-raw-material). However, similar to the CV variation for the Phase 2, the results of site variation show that CV of body length, maximum length, mid width, and tang width are generally lower than tang length, stem length, and stem width, indicating the combination of the two biases. The attributes with lower CV values could be taught through indirect bias while the higher CV attributes could be depending more on various manufacturing situations. We assume that the lower CV values of the four attributes could imply that those four attributes are closely associated with the function of the tool so that the knappers wanted to standardize the shape. For example, body length is related to penetration, durability, hardness, and rejuvenation of a projectile point and tang width is connected to wound damage, penetration, durability and hardness [@odell1986experiments; @shea2001experimental; @cheshier2006projectile; @yaroshevich2016projectile; @bebber2017developing; @wood2018wound]. This combination model describes that there were social connections between groups that share the tool production technique. Individuals or groups learned the major technology of stemmed points by copying a model from another individual or group (indirect bias). Then, those individuals or groups adjusted their technique and shape of the stem depending on their needs and material availability. Our results recalls -@lee2013current's model but his model explains combination of the *in situ* tradition for flake tools and introduction of blade-associate toolkit including stemmed points. On the other hand, we saw the combination of the *in situ* tradition and introduced stemmed point technology within the single artifact.

>>>>

Following @bettinger1999point we computed Pearson's correlation coefficients for all measurements of the seven attributes that we measured on the stemmed points. The correlation of artefact attributes was proposed by @bettinger1999point to be a key indicator of indirect bias during cultural transmission. This is based on the assumption that assumed that artefact-making was transmitted in packages of traits taken from socially successful individuals who were role-models. In their work points from Eastern California and Central Nevada, @bettinger1999point interpreted correlation coefficient values of around 0.5 and higher as evidence of indirect bias, and lower values as evidence of guided variation.

@@

Panel A of @fig-plot-cv-all-attributes shows a relatively wide distribution of body length (BL) and maximum length (ML) while the other attributes are more similar and shorter (under 2.5 cm). Among these attributes, stem width (SW) is the smallest, regardless of overall size of the stemmed points.

@@

<!-- GP to write up the results in detail, that you see in the plot made by the code below. Use this paper as a template for how detailed you need to be here: "Point Typologies, Cultural Transmission, and the Spread of Bow-and-Arrow Technology" -->

@@

To investigate the social contexts of technological transitions represented by the emergence of stemmed points during the Korean Late Paleolithic, we applied concepts of cultural transmission. We asked three questions to examine the dominant mode of cultural transmission and the transmission process over time: what is the dominant mode of cultural transmission for technological innovation in the Korean Late Paleolithic? Do the modes of cultural transmission vary over time? and Do the modes of cultural transmission vary over space? We focused on two transmission biases: guided variation (socially isolated groups made stemmed points through trial and error) and indirect bias (socially connected groups whose knowledge of stemmed points derived from copying others). We set roughly `r threshold_value` as our threshold of the CV value to decide between these two types of transmission biases, based on prior work on Great Basin Projectiles [@eerkens2001techniques]. If the CV value is lower than 25, we interpreted this as indicating indirect bias while values higher than 25 were interpreted as guided variation.

The CV values for all stemmed point attributes are mostly over 25, except for maximum length, which is just below. Change in CV values for over time in the Korean Late Paleolithic period is complex, with no clear directional changes (@fig-plot-cv-over-time). Comparing chronological phases 2 and 3, we found that there are only minor differences between the two phases. Applying the Modified signed-likelihood ratio test (SLRT) to test for the equality of CVs, we found no statistically significant changes in the CV values of any attributes (@tbl-cv-over-time) between the two phases. In exploring individual assemblages that contain multiple stemmed points we found confirmation of minimal change in CV values over time (@fig-plot-cv-four-assemblage).

@@

In exploring geographical variation we found no clear pattern variation in CV between sites with more than five stemmed points (@fig-plot-cv-by-site). The most striking pattern in this site comparison is that CV values of body length, maximum length, mid width, and tang width are generally lower than tang length, stem length, and stem width, similar to what is evident in @fig-plot-cv-all-attributes. The attributes with lower CV values could be transmitted in a social context dominated by indirect bias, while the higher CV attributes could be influenced more by specific manufacturing and maintenance situations. We assume that the lower CV values of the four attributes could imply that those four attributes are closely associated with the projectile function of the tool so that the knappers wanted to standardize the shape. For example, body length is related to penetration, durability, hardness, and rejuvenation of a projectile point and tang width is connected to wound damage, penetration, durability and hardness [@odell1986experiments; @shea2001experimental; @cheshier2006projectile; @yaroshevich2016projectile; @bebber2017developing; @wood2018wound]. This can be contrasted with tang length, stem length, and stem width, which are away from the impact end of the artefact. These were allowed to vary more freely between points, perhaps to be able to accommodate shafts of different types of wood with varying properties of strength, flexibility, and weight.

The most striking differences in CV values is evident in comparisons of points grouped by raw materials. Artefacts made from quartzite have much higher CV values, and wider confidence intervals, than all other raw materials (@fig-plot-cv-by-raw-material). This likely reflects the relatively unpredictable knapping quality of quartzite due to the presence of visible grains. These grains can redirect knapping forces in unexpected directions, resulting in unintended flake removals. This makes is harder for the knapper to repeatedly produce the same size and shape end product. This hints at the possibility that physical properties of the stone artefact raw materials were a key mechanism in determining shape and size variability. However with only three artefacts made from quartzite in our sample, further analysis is needed to make robust conclusions about the influence of raw materials here.

However, similar to the CV variation for the Phase 2,  This combination model describes that there were social connections between groups that share the tool production technique. Individuals or groups learned the major technology of stemmed points by copying a model from another individual or group (indirect bias). Then, those individuals or groups adjusted their technique and shape of the stem depending on their needs and material availability. Our results recalls -@lee2013current's model but his model explains combination of the *in situ* tradition for flake tools and introduction of blade-associate toolkit including stemmed points. On the other hand, we saw the combination of the *in situ* tradition and introduced stemmed point technology within the single artifact.
